\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-15 1-intro.Rmd
merge general information of spatio-temporal data  in intro and section 2

Spatio-temporal data concerns variables in the spatial coordinates across time. Some variables are invariant to time and can be solely identified by its spatial location while other time varying variables need both the spatial and temporal identifiers to identify. Manipulating data in these two levels is usually done separately and the third tidy data principal [@tidydata] recommends each type of observational unit to form a table. Another feature for a spatio-temporal data structure to have is that variables created from subsequent operations can be automatically added to the existing table with the same level. This is better than an output that index the variables with the identifiers since otherwise analysts will need to manually join this output with existing variables in the same level.

@@

Currently, available spatio-temporal data structures in R includes: \pkg{spacetime} [@spacetime], which proposes four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). \pkg{spacetime} is built on \pkg{sp} [@sp] and \pkg{xts} [@xts], both of which have been replaced by more recent implementations. \pkg{spatstat} [@spatstat] implements a `ppp` class for point pattern data; and more recent, \pkg{stars} [@stars] implements a spatio-temporal for vector and raster spatio-temporal data.  While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables at each temporal unit; or split them into two separate tables that analysts need to manually join the data constantly when doing the analysis. This creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any variables that are measured at fixed location across time.

>>>>

Spatio-temporal data concerns variables in the space across time. \textcolor{red}{Elaborate on variable, space, and time}. If the time interval is regular and each site has fixed point location, it can also be thought of as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubical layout, variables can be those that are invariant to time, i.e. the longitude and latitude coordinates, or those that varying across time. While conceptually the data format is relatively simple, the actual data object that comes to the analysts can be in various forms in practice. Figure \ref{fig:illu-input} shows four examples where the first one combines the time invariant and varying variables into a single table through duplicating the time invariant variables; 2) splits those variables into two separate sheets; 3) ...; and lastly 4) considers the array-based data, like NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others,  to organise multivariate data in a spatio-temporal grid.

\textcolor{red}{think through a cube of long x lat x time vs. id x time x variable}

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data format: (1) a single table for all the variables; (2) separate tables for spatial and temporal variables; (3) a table of metadata used to query the database and a separate table for queried data; and (4) an array data structure of cube or hypercube.", fig.align='center', out.height = "40%"}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

<!-- Manipulating data in these two levels is usually done separately and the third tidy data principal [@tidydata] recommends each type of observational unit to form a table. Another feature for a spatio-temporal data structure to have is that variables created from subsequent operations can be automatically added to the existing table with the same level. This is better than an output that index the variables with the identifiers since otherwise analysts will need to manually join this output with existing variables in the same level. -->

@@

\textcolor{red}{more beginning like "working with such data different authors ..."} Currently, there's various different spatio-temporal data structures in R includes: \pkg{spacetime} [@spacetime], which proposes four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). \pkg{spacetime} is built on \pkg{sp} [@sp] and \pkg{xts} [@xts], both of which have been replaced by more recent implementations. \pkg{spatstat} [@spatstat] implements a `ppp` class for point pattern data; and more recent, \pkg{stars} [@stars] implements a spatio-temporal for vector and raster spatio-temporal data.

In spatio-temporal data analysis, analysts will receive one of the various data formats depends on the data provider and need to manually rearrange the data to fit into one of the data object in \proglang{R} before starting any analysis on the data itself. Looking at spatial or temporal data analysis, data objects like \pkg{sf} [@sf] and \pkg{tsibble} [@tsibble] have smoothed the data analysis in these two domains while there has not yet been a spatio-temporal data structure that makes it easy to explore spatio-temporal data and this creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any variables that are measured at fixed location across time.

<!-- While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables at each temporal unit; or split them into two separate tables that analysts need to manually join the data constantly when doing the analysis. \textcolor{red}{There has been data structure proposed for spatial (\pkg{sf} by @sf) and temporal (\pkg{tsibble} by @tsibble) data in the R community, while less has been done for spatio-temporal data.}  -->

### 2022-02-15 2-the-cubble-pkg.Rmd
a few small changes as per writing the vignette
merge general information of spatio-temporal data  in intro and section 2

Spatio-temporal data usually come to the analysts in various forms and Figure \ref{fig:illu-input} shows four examples of this. The first three are variations of tabular data that uses a single table (1) or multiple tables (2 & 3) while the fourth uses a cube or hyper-cube to to represent the data. An example of the fourth type is the NetCDF (Network Common Data Form), commonly used in the climatology community to deliver global mapping of atmosphere, ocean, and land.

These four examples show the common characters of the spatio-temporal data: 1) a spatial identifier that uniquely identifies each site, 2) a temporal identifier that prescribes the time stamp each site is recorded, 3) Coordinates, comprising of latitude and longitude, that locate the site on the map. These identifiers will be the building blocks for cubble. Another way to structure the spatio-temporal data is to think of it as a data cube with three axes: id, time, and variable. Notice that in the time-variable face of this cube, time invariant variables would share the same value across all the time stamps, while time varying variables have values varying across time. (Maybe add a cube diagram here). This heterogeneity will be used to split the data into two forms in cubble.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data format: (1) a single table for all the variables; (2) separate tables for spatial and temporal variables; (3) a table of metadata used to query the database and a separate table for queried data; and (4) an array data structure of cube or hypercube.", fig.align='center', out.height = "40%"}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

In a cubble, there are two forms: nested form and long form, and Figure \ref{fig:illu-cubble} sketches the two forms with the associated attributes. The decision on which form to use in data manipulation is output-oriented, meaning analysts need to first think about whether the output of an operation is identified only by the spatial identifier, or a combination of spatial and temporal identifier. The nested cubble is suitable for working with operations that are only identified by the site and this type of operation can be a pure manipulation of time invariant variables, or an operation that summarises time varying variables into site. Underneath the nested form, a cubble is built from a \code{rowwise_df} data frame where each site forms a separate group. This structure simplifies the calculation that involves temporal variables by avoiding the use of \code{purrr::map()} syntax when working with list-column.

@@

## Create a cubble in the nested form
A cubble can be created with \code{as_cubble()} by supplying the three key components: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order of longitude and latitude. The naming of \code{key} and \code{index} follows the convention in the \pkg{tsibble} package. Below is an example of creating a cubble and by default, the cubble created is in the nested form:

@@

@@

Cubble is also compatible with \pkg{tsibble}. When creating a cubble from a tsibble, Only the \code{coords} argument needs to be specified:

>>>>

\textcolor{red}{more beginning} In cubble data can represented in two formats: nested form and long form, and Figure \ref{fig:illu-cubble} sketches the two forms with the associated attributes. The decision on which form to use in data manipulation is output-oriented, meaning analysts need to first think about whether the output of an operation is identified only by the spatial identifier, or a combination of spatial and temporal identifier. The nested cubble is suitable for working with operations that are only identified by the site and this type of operation can be a pure manipulation of time invariant variables, or an operation that summarises time varying variables into site. Underneath the nested form, a cubble is built from a \code{rowwise_df} data frame where each site forms a separate group. This structure simplifies the calculation that involves temporal variables by avoiding the use of \code{purrr::map()} syntax when working with list-column.

@@

## Create a cubble in the nested form

\textcolor{red}{more beginning} A cubble can be created with \code{as_cubble()} by supplying the three key components: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order of longitude and latitude. The naming of \code{key} and \code{index} follows the convention in the \pkg{tsibble} package. \textcolor{red}{give some context of the data about to be used: a subset of a larger data etc} Below is an example of creating a cubble and by default, the cubble created is in the nested form:

@@

\textcolor{red}{A beginning sentence that summarises the subsection}. 
\textcolor{red}{Given more details and examples when going through each subsection}

@@

Cubble is also compatible with \pkg{tsibble}. When creating a cubble from a tsibble, only the \code{coords} argument needs to be specified:

### 2022-02-15 3-advanced.Rmd
clean up bibbib
merge general information of spatio-temporal data  in intro and section 2

Cubble provides functionality for matching sites from two data sources. To be a valid pair of matching, the matched pair from different data sources need to:

@@

This functionality can be seen as a matching exercise [@stuart2010matching; @mcintosh2018using] in the spatio-temporal domain to pair up nearby observations. It can also be considered as a form of data fusion [@castanedo2013review; @cocchi2019data], where data from multiple sources are combined together.

>>>>

\textcolor{red}{more beginning} Cubble provides functionality for matching sites from two data sources. To be a valid pair of matching, the matched pair from different data sources need to:

@@

This functionality can be seen as a matching exercise [@stuart2010matching; @mcintosh2018using] in the spatio-temporal domain to pair up nearby observations. It can also be considered as a form of data fusion [@cocchi2019data], where data from multiple sources are combined together.

### 2022-02-15 main.tex
merge general information of spatio-temporal data  in intro and section 2

Spatio-temporal data concerns variables in the spatial coordinates
across time. Some variables are invariant to time and can be solely
identified by its spatial location while other time varying variables
need both the spatial and temporal identifiers to identify. Manipulating
data in these two levels is usually done separately and the third tidy
data principal \citep{tidydata} recommends each type of observational
unit to form a table. Another feature for a spatio-temporal data
structure to have is that variables created from subsequent operations
can be automatically added to the existing table with the same level.
This is better than an output that index the variables with the
identifiers since otherwise analysts will need to manually join this
output with existing variables in the same level.

Currently, available spatio-temporal data structures in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). \pkg{spacetime} is built on \pkg{sp} \citep{sp} and
\pkg{xts} \citep{xts}, both of which have been replaced by more recent
implementations. \pkg{spatstat} \citep{spatstat} implements a
\texttt{ppp} class for point pattern data; and more recent, \pkg{stars}
\citep{stars} implements a spatio-temporal for vector and raster
spatio-temporal data. While these implementations either store spatial
and temporal variables all in a single table, hence duplicate the
spatial variables at each temporal unit; or split them into two separate
tables that analysts need to manually join the data constantly when
doing the analysis. This creates a gap in the software development. The
requirement for such a tool is important given the ubiquity of
spatio-temporal vector data in the wild: climate observations from
weather stations, air quality variables from air monitoring stations, or
any variables that are measured at fixed location across time.

Spatio-temporal data usually come to the analysts in various forms and
Figure \ref{fig:illu-input} shows four examples of this. The first three
are variations of tabular data that uses a single table (1) or multiple
tables (2 \& 3) while the fourth uses a cube or hyper-cube to to
represent the data. An example of the fourth type is the NetCDF (Network
Common Data Form), commonly used in the climatology community to deliver
global mapping of atmosphere, ocean, and land.

These four examples show the common characters of the spatio-temporal
data: 1) a spatial identifier that uniquely identifies each site, 2) a
temporal identifier that prescribes the time stamp each site is
recorded, 3) Coordinates, comprising of latitude and longitude, that
locate the site on the map. These identifiers will be the building
blocks for cubble. Another way to structure the spatio-temporal data is
to think of it as a data cube with three axes: id, time, and variable.
Notice that in the time-variable face of this cube, time invariant
variables would share the same value across all the time stamps, while
time varying variables have values varying across time. (Maybe add a
cube diagram here). This heterogeneity will be used to split the data
into two forms in cubble.

@@

In a cubble, there are two forms: nested form and long form, and Figure
\ref{fig:illu-cubble} sketches the two forms with the associated
attributes. The decision on which form to use in data manipulation is
output-oriented, meaning analysts need to first think about whether the
output of an operation is identified only by the spatial identifier, or
a combination of spatial and temporal identifier. The nested cubble is
suitable for working with operations that are only identified by the
site and this type of operation can be a pure manipulation of time
invariant variables, or an operation that summarises time varying
variables into site. Underneath the nested form, a cubble is built from
a \code{rowwise_df} data frame where each site forms a separate group.
This structure simplifies the calculation that involves temporal
variables by avoiding the use of \code{purrr::map()} syntax when working
with list-column.

@@

A cubble can be created with \code{as_cubble()} by supplying the three
key components: \code{key} as the spatial identifier; \code{index} as
the temporal identifier; and a vector of \code{coords} in the order of
longitude and latitude. The naming of \code{key} and \code{index}
follows the convention in the \pkg{tsibble} package. Below is an example
of creating a cubble and by default, the cubble created is in the nested
form:

>>>>

Spatio-temporal data concerns variables in the space across time.
\textcolor{red}{Elaborate on variable, space, and time}. If the time
interval is regular and each site has fixed point location, it can also
be thought of as a data cube with three axes: spatial identifier,
temporal identifier, and variable. In this cubical layout, variables can
be those that are invariant to time, i.e.~the longitude and latitude
coordinates, or those that varying across time. While conceptually the
data format is relatively simple, the actual data object that comes to
the analysts can be in various forms in practice. Figure
\ref{fig:illu-input} shows four examples where the first one combines
the time invariant and varying variables into a single table through
duplicating the time invariant variables; 2) splits those variables into
two separate sheets; 3) \ldots; and lastly 4) considers the array-based
data, like NetCDF (Network Common Data Form) data, which is commonly
used in earth science, e.g.~climatology, meteorology, and oceanography,
among others, to organise multivariate data in a spatio-temporal grid.

\textcolor{red}{think through a cube of long x lat x time vs. id x time x variable}

@@

\textcolor{red}{more beginning like "working with such data different authors ..."}
Currently, there's various different spatio-temporal data structures in
R includes: \pkg{spacetime} \citep{spacetime}, which proposes four
space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI),
and trajectory (STT). \pkg{spacetime} is built on \pkg{sp} \citep{sp}
and \pkg{xts} \citep{xts}, both of which have been replaced by more
recent implementations. \pkg{spatstat} \citep{spatstat} implements a
\texttt{ppp} class for point pattern data; and more recent, \pkg{stars}
\citep{stars} implements a spatio-temporal for vector and raster
spatio-temporal data.

In spatio-temporal data analysis, analysts will receive one of the
various data formats depends on the data provider and need to manually
rearrange the data to fit into one of the data object in \proglang{R}
before starting any analysis on the data itself. Looking at spatial or
temporal data analysis, data objects like \pkg{sf} \citep{sf} and
\pkg{tsibble} \citep{tsibble} have smoothed the data analysis in these
two domains while there has not yet been a spatio-temporal data
structure that makes it easy to explore spatio-temporal data and this
creates a gap in the software development. The requirement for such a
tool is important given the ubiquity of spatio-temporal vector data in
the wild: climate observations from weather stations, air quality
variables from air monitoring stations, or any variables that are
measured at fixed location across time.

\textcolor{red}{more beginning} In cubble data can represented in two
formats: nested form and long form, and Figure \ref{fig:illu-cubble}
sketches the two forms with the associated attributes. The decision on
which form to use in data manipulation is output-oriented, meaning
analysts need to first think about whether the output of an operation is
identified only by the spatial identifier, or a combination of spatial
and temporal identifier. The nested cubble is suitable for working with
operations that are only identified by the site and this type of
operation can be a pure manipulation of time invariant variables, or an
operation that summarises time varying variables into site. Underneath
the nested form, a cubble is built from a \code{rowwise_df} data frame
where each site forms a separate group. This structure simplifies the
calculation that involves temporal variables by avoiding the use of
\code{purrr::map()} syntax when working with list-column.

@@

\textcolor{red}{more beginning} A cubble can be created with
\code{as_cubble()} by supplying the three key components: \code{key} as
the spatial identifier; \code{index} as the temporal identifier; and a
vector of \code{coords} in the order of longitude and latitude. The
naming of \code{key} and \code{index} follows the convention in the
\pkg{tsibble} package.
\textcolor{red}{give some context of the data about to be used: a subset of a larger data etc}
Below is an example of creating a cubble and by default, the cubble
created is in the nested form:
